On the reality of the quantum state once again

### 2022-01-22 Articles/2021-HarriganSpekkens/HarriganSpekkens.tex
HarriganSpekkens: another pass, mainly minor typos

\documentclass[twocolumn,prl,floatfix,superscriptaddress]{revtex4-1}
\usepackage{graphicx, subfigure, amsfonts, amssymb, amsmath, array, gensymb, fontenc, color, lipsum}
%\documentclass[aps,twocolumn,showpacs,preprintnumbers]{revtex4}
%\usepackage{graphicx}  % Include figure files
%\usepackage{subfigure}
%\usepackage{multirow}
%\linespread{1.0}
%\usepackage{fancyhdr}
%\usepackage{longtable}
%\usepackage{parskip}
%\usepackage[T1]{fontenc}
%\usepackage{dcolumn}
%\usepackage{bm}        
%\usepackage{amsfonts}  
%\usepackage{amsmath}   
%\usepackage{amssymb}
\usepackage{hyperref}
\hypersetup{colorlinks= true, allcolors=blue}
%\setcitestyle{aysep={}}
%\setlength{\parindent}{11pt}
\frenchspacing

@@

\title{On the reality of the quantum state once again}
\author{Gabriele Carcassi}
\affiliation{Physics Department, University of Michigan, Ann Arbor, MI 48109}
\author{Andrea Oldofredi}
\affiliation{Centre of Philosophy (CFUL), University of Lisbon, Portugal}
\author{Christine Aidala}
\affiliation{Physics Department, University of Michigan, Ann Arbor, MI 48109}
\vspace{2mm}

@@

In 2012 Pusey, Barrett and Rudolph published a formal result in \emph{Nature Physics}, widely known as the PBR theorem, showing that ``if the quantum state merely represents information about the real physical state of a system, then experimental predictions are obtained that contradict those of quantum theory'' (\cite{PBR:2012}, p.\ 475). Alternatively stated, PBR argued that in every model reproducing the statistics and predictions of Quantum Mechanics (QM) the quantum state $\psi$ must represent real physical properties of the system under consideration and not agents' knowledge---i.e.\ models must be $\psi-$ontic. Consequently, quantum theories cannot be $\psi-$epistemic.

@@

In this paper we are not going to rebut the theorem itself. We instead aim to carefully reexamine one of its premises, i.e.\ Harrigan and Spekkens' (HS) classification between $\psi-$ontic and $\psi-$epistemic ontological models  \cite{Harrigan:2010} on top of which the PBR result is derived. We believe that a careful examination of this assumption forces us to reevaluate the significance of the PBR conclusion.

HS proposed a rigorous classification in order to categorize the nature of the quantum state, i.e.\ to establish whether in a certain model $\psi$ corresponds to a real property of a quantum object, in which case the model is called $\psi-$ontic, or to some observer information, making it $\psi-$epistemic. While the original aim of this classification was to clarify Einstein's view of quantum theory, the HS framework has been widely employed in the literature not only to categorize different interpretations, but also to argue what types of interpretations are admissible (\cite{Leifer:2013, Leifer:2017, Branciard:2014, Hermens:2021, Wood:2015, Ringbauer:2015, Mazurek:2016, Bartlett:2012}; cf.\ \cite{Oldofredi:2020b, Lewis:2012, Ladyman:2021} for critical discussions). Given the influence that the HS framework and the PBR theorem have in quantum foundations, it is worthwhile to look again at the former with more scrutiny in order to understand whether it is a sound basis from which to draw conclusions on the nature of the quantum state.

@@

\begin{figure}
\includegraphics[scale=.7]{ontic}
\caption{\footnotesize{Harrigan and Spekkens distinction between $\psi-$ontic (a) and $\psi-$epistemic (b) ontological models.}}\label{overlap}
\end{figure}

@@

Let us first characterize the extent of the discrepancy. In HS, the space of epistemic states $E_{HS}$ is the set of probability distributions over pure states, which is an infinite dimensional space. This is a direct consequence of defining the epistemic states as a distribution $p(\lambda|P)$ that, for each preparation $P$, assigns a well defined probability to each ontological state $\lambda$ \footnote{We will follow the convention of not distinguishing between probability and probability density.}. In QM, mixed states $E_{Q}$ are in one-to-one correspondence with the interior of the Bloch sphere, a three dimensional manifold. Looking at the difference in dimension, the discrepancy is not minor.

Note that the space of pure states $\Lambda$ is the same in both cases, one state for each direction of spin. Also note that all epistemic states in both spaces can be reached by mixture of pure states. The difference is that, in the quantum case, we have an ``aliasing'': different mixtures will give the same mixed states. Since each classical probability distribution corresponds to a unique mixing of pure states, we have a map $\iota : E_{HS} \mapsto E_{Q}$ that allows us to go from HS-epistemic states to density matrices, the Q-epistemic states. This map is surjective but not injective. Therefore $E_{Q}$ can be seen as a set of equivalence classes of $E_{HS}$. The key question, then, is the following: is the map $\iota$ sufficiently well behaved that we can understand the epistemic content of $E_{Q}$ based on the epistemic content of $E_{HS}$? Or is it so ill behaved that there is no connection between the two?

Ideally, one would like to say that $E_{HS}$ represents all possible epistemic states that include the knowledge about preparation, while $E_{Q}$ represents only those distinguishable through measurement. For example, one may want to say that the axis of preparation is, in principle, an ontologically distinct property, but this is not experimentally distinguishable. Therefore, distinct HS-epistemic states for each direction are projected by $\iota$ onto the same Q-epistemic state of maximum uncertainty. This sounds very plausible and appealing, but it turns out this is not the case. To show this fact we will concentrate on a single aspect of the map: how information entropy transforms under $\iota$. While there are other problems, we do not need to explore them: given the crucial role of entropy in information theory, its breakdown is sufficient to show that the epistemic structures of $E_{HS}$ and $E_{Q}$ are different.

Guided by figure \ref{fig_map}, let us study the map $\iota$ by going through a few examples. Since every pure state in quantum mechanics has zero entropy, we will require the same for all pure states in $E_{HS}$. Let us consider the HS-epistemic states specified by three uniform distributions over the Bloch sphere with support given by:
\begin{description}
	\item[A] the poles, $z^+$ and $z^-$;
	\item[B] the equator, all states on the x/y planes;
	\item[C] the whole sphere, all points.
\end{description}
Since we want the entropy for a single pure state to be zero, the entropy for case \textbf{A} will be 1 (uniform distribution between two discrete cases). Since $\iota(A)$ will be the maximally mixed state, this also corresponds to entropy 1. So far so good: $\iota$ maps entropy 1 to entropy 1! However, note that \textbf{A} is a measure zero set of \textbf{B}, which is a measure zero set of \textbf{C}. The entropy of \textbf{C} is therefore infinite with respect to \textbf{B} which is infinite with respect to \textbf{A}. On the other hand, \textbf{A}, \textbf{B} and \textbf{C} will all correspond to the same maximally mixed state, therefore $\iota$ collapses together epistemic states at different levels of infinity. Note that, in this case, $\iota$ picks the minimum entropy between the equivalent epistemic states. This is inconsistent with the idea that $E_{Q}$ represents states where knowledge is lost: that would pick the maximum possible entropy.

\begin{figure}
\includegraphics[scale=.4]{fig2}
\caption{\footnotesize{The function $\iota$ maps the epistemic states $E_{HS}$ of a $\psi$-ontic model of a single qubit to the respective density matrices $E_Q$. Note that neither the value nor the ordering of the information entropy is preserved.}}\label{fig_map}
\end{figure}

Now consider another uniform distribution over the support:
\begin{description}
	\item[D] the north pole and a point on the equator, $z^+$ and $x^+$.
\end{description}
This is yet another uniform distribution between two discrete cases, therefore in $E_{HS}$ the entropy is 1. However, $\iota(D)$ is not the maximally mixed state and therefore the entropy is less than 1. This means that epistemic states that are isoentropic in $E_{HS}$ are not isoentropic in $E_{Q}$. If we think as $E_{HS}$ and $E_{Q}$ as partially ordered by information entropy, then $\iota$ is not an order preserving map. Therefore, the epistemic structures represented by $E_{HS}$ and $E_{Q}$ are not isomorphic.

@@

In this essay we showed not only that HS framework does not correctly describe the epistemic structure of QM, but also that it implicitly assumes non-contextuality. Hence, it cannot be employed to draw conclusions on the nature of quantum states. Consequently, any other result relying on HS categorization inherits the same issues affecting it. In particular, it follows that the PBR theorem is derived from a technically ill-defined basis; thus, it is demonstrated neither that quantum states must necessarily be HS-$\psi-$ontic, nor that epistemic interpretations are contradicted by quantum theory. Concluding, we believe that while work in quantum foundations typically focuses on measurements and their effect on pure states, equal importance must be given to preparations and their effect on mixed states. The different behavior of entropy in classical and quantum mechanics has profound implications not just for information theory, but also for statistical mechanics and its measurable predictions.

@@

Andrea Oldofredi is grateful to the Funda\c{c}$\tilde{\mathrm{a}}$o para a Ci\^encia e a Tecnologia (FCT) for financial support (Grant no. 2020.02858).

%\bibliographystyle{ieeetr}
\bibliography{PhDthesis}

>>>>

\documentclass[twocolumn,prl,floatfix,superscriptaddress]{revtex4-2}
\usepackage{graphicx, subfigure, amsfonts, amssymb, amsmath, array, gensymb, fontenc, color, lipsum}
%\documentclass[aps,twocolumn,showpacs,preprintnumbers]{revtex4}
%\usepackage{graphicx}  % Include figure files
%\usepackage{subfigure}
%\usepackage{multirow}
%\linespread{1.0}
%\usepackage{fancyhdr}
%\usepackage{longtable}
%\usepackage{parskip}
%\usepackage[T1]{fontenc}
%\usepackage{dcolumn}
%\usepackage{bm}        
%\usepackage{amsfonts}  
%\usepackage{amsmath}   
%\usepackage{amssymb}
\usepackage{hyperref}
\hypersetup{colorlinks= true, allcolors=blue}
%\setcitestyle{aysep={}}
%\setlength{\parindent}{11pt}
\frenchspacing

@@

\title{On the reality of the quantum state once again}
\author{Gabriele Carcassi}
\affiliation{Physics Department, University of Michigan, Ann Arbor, MI 48109}
\author{Andrea Oldofredi}
\affiliation{Centre of Philosophy (CFUL), University of Lisbon, Portugal}
\author{Christine A. Aidala}
\affiliation{Physics Department, University of Michigan, Ann Arbor, MI 48109}
\vspace{2mm}

@@

In 2012 Pusey, Barrett and Rudolph published a formal result in \emph{Nature Physics}, widely known as the PBR theorem, showing that ``if the quantum state merely represents information about the real physical state of a system, then experimental predictions are obtained that contradict those of quantum theory'' (\cite{PBR:2012}, p.\ 475). Alternatively stated, PBR argued that in every model reproducing the statistics and predictions of Quantum Mechanics (QM) the quantum state $\psi$ must represent real physical properties of the system under consideration and not agents' knowledge---i.e.\ models must be $\psi$-ontic. Consequently, quantum theories cannot be $\psi$-epistemic.

@@

In this paper we are not going to rebut the theorem itself. We instead aim to carefully reexamine one of its premises, i.e.\ Harrigan and Spekkens' (HS) classification between $\psi$-ontic and $\psi$-epistemic ontological models  \cite{Harrigan:2010} on top of which the PBR result is derived. We believe that a careful examination of this assumption forces us to reevaluate the significance of the PBR conclusion.

HS proposed a rigorous classification in order to categorize the nature of the quantum state, i.e.\ to establish whether in a certain model $\psi$ corresponds to a real property of a quantum object, in which case the model is called $\psi$-ontic, or to some observer information, making it $\psi$-epistemic. While the original aim of this classification was to clarify Einstein's view of quantum theory, the HS framework has been widely employed in the literature not only to categorize different interpretations, but also to argue what types of interpretations are admissible (\cite{Leifer:2013, Leifer:2017, Branciard:2014, Hermens:2021, Wood:2015, Ringbauer:2015, Mazurek:2016, Bartlett:2012}; cf.\ \cite{Oldofredi:2020b, Lewis:2012, Ladyman:2021} for critical discussions). Given the influence that the HS framework and the PBR theorem have in quantum foundations, it is worthwhile to look again at the former with more scrutiny in order to understand whether it is a sound basis from which to draw conclusions on the nature of the quantum state.

@@

\begin{figure}
\includegraphics[scale=.7]{ontic}
\caption{\footnotesize{Harrigan and Spekkens' distinction between $\psi$-ontic (a) and $\psi$-epistemic (b) ontological models.}}\label{overlap}
\end{figure}

@@

Let us first characterize the extent of the discrepancy. In HS, the space of epistemic states $E_{HS}$ is the set of probability distributions over pure states, which is an infinite dimensional space. This is a direct consequence of defining the epistemic states as a distribution $p(\lambda|P)$ that, for each preparation $P$, assigns a well defined probability to each ontological state $\lambda$ \footnote{We will follow the convention of not distinguishing between probability and probability density.}. In QM, mixed states $E_{QM}$ are in one-to-one correspondence with the interior of the Bloch sphere, a three dimensional manifold. Looking at the difference in dimension, the discrepancy is not minor.

Note that the space of pure states $\Lambda$ is the same in both cases, one state for each direction of spin. Also note that all epistemic states in both spaces can be reached by mixture of pure states. The difference is that, in the quantum case, we have an ``aliasing'': different mixtures will give the same mixed states. Since each classical probability distribution corresponds to a unique mixing of pure states, we have a map $\iota : E_{HS} \mapsto E_{QM}$ that allows us to go from HS-epistemic states to density matrices, the QM-epistemic states. This map is surjective but not injective. Therefore $E_{QM}$ can be seen as a set of equivalence classes of $E_{HS}$. The key question, then, is the following: is the map $\iota$ sufficiently well behaved that we can understand the epistemic content of $E_{QM}$ based on the epistemic content of $E_{HS}$? Or is it so ill behaved that there is no connection between the two?

Ideally, one would like to say that $E_{HS}$ represents all possible epistemic states that include the knowledge about preparation, while $E_{QM}$ represents only those distinguishable through measurement. For example, one may want to say that the axis of preparation is, in principle, an ontologically distinct property, but this is not experimentally distinguishable. Therefore, distinct HS-epistemic states for each direction are projected by $\iota$ onto the same QM-epistemic state of maximum uncertainty. This sounds very plausible and appealing, but it turns out this is not the case. To show this fact we will concentrate on a single aspect of the map: how information entropy transforms under $\iota$. While there are other problems, we do not need to explore them: given the crucial role of entropy in information theory, its breakdown is sufficient to show that the epistemic structures of $E_{HS}$ and $E_{QM}$ are different.

Guided by figure \ref{fig_map}, let us study the map $\iota$ by going through a few examples. Since every pure state in quantum mechanics has zero entropy, we will require the same for all pure states in $E_{HS}$. Let us consider the HS-epistemic states specified by three uniform distributions over the Bloch sphere with support given by:
\begin{description}
	\item[A] the poles, $z^+$ and $z^-$;
	\item[B] the equator, all states on the x/y planes;
	\item[C] the whole sphere, all points.
\end{description}
Since we want the entropy for a single pure state to be zero, the entropy for case \textbf{A} will be 1 (uniform distribution between two discrete cases). Since $\iota(A)$ will be the maximally mixed state, this also corresponds to entropy 1. So far so good: $\iota$ maps entropy 1 to entropy 1. However, note that \textbf{A} is a measure zero set of \textbf{B}, which is a measure zero set of \textbf{C}. The entropy of \textbf{C} is therefore infinite with respect to \textbf{B} which is infinite with respect to \textbf{A}. On the other hand, \textbf{A}, \textbf{B} and \textbf{C} will all correspond to the same maximally mixed state, therefore $\iota$ collapses together epistemic states at different levels of infinity. Note that, in this case, $\iota$ picks the minimum entropy between the equivalent epistemic states. This is inconsistent with the idea that $E_{QM}$ represents states where knowledge is lost: that would pick the maximum possible entropy.

\begin{figure}
\includegraphics[scale=.4]{fig2}
\caption{\footnotesize{The function $\iota$ maps the epistemic states $E_{HS}$ of a $\psi$-ontic model of a single qubit to the respective density matrices $E_{QM}$. Note that neither the value nor the ordering of the information entropy is preserved.}}\label{fig_map}
\end{figure}

Now consider another uniform distribution over the support:
\begin{description}
	\item[D] the north pole and a point on the equator, $z^+$ and $x^+$.
\end{description}
This is yet another uniform distribution between two discrete cases, therefore in $E_{HS}$ the entropy is 1. However, $\iota(D)$ is not the maximally mixed state and therefore the entropy is less than 1. This means that epistemic states that are isoentropic in $E_{HS}$ are not isoentropic in $E_{QM}$. If we think of $E_{HS}$ and $E_{QM}$ as partially ordered by information entropy, then $\iota$ is not an order preserving map. Therefore, the epistemic structures represented by $E_{HS}$ and $E_{QM}$ are not isomorphic.

@@

We showed not only that the HS framework does not correctly describe the epistemic structure of QM, but also that it implicitly assumes non-contextuality. Hence, it cannot be employed to draw conclusions on the nature of quantum states. Consequently, any other result relying on the HS categorization inherits the same issues. In particular, it follows that the PBR theorem is derived from a technically ill-defined basis; thus, it is demonstrated neither that quantum states must necessarily be HS-$\psi$-ontic, nor that epistemic interpretations are contradicted by quantum theory. Concluding, we believe that while work in quantum foundations typically focuses on measurements and their effect on pure states, equal importance must be given to preparations and their effect on mixed states. The different behavior of entropy in classical and quantum mechanics has profound implications not just for information theory, but also for statistical mechanics and its measurable predictions.

@@

Andrea Oldofredi is grateful to the Funda\c{c}$\tilde{\mathrm{a}}$o para a Ci\^encia e a Tecnologia (FCT) for financial support (Grant no. 2020.02858).  This work is in connection to Assumptions of Physics, a larger project that aims to identify a handful of physical principles from which the basic laws can be rigorously derived  (\url{https://assumptionsofphysics.org}).

%\bibliographystyle{ieeetr}
\bibliography{HarriganSpekkens}
